Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-30 text/front_end_v3.qmd
finished editing front end

**Geomarket Filter in Student Search Service**. By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the “enrollment funnel.” @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.[^11] Historically, dominant list vendors are college board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

@@

However, @RN5012 argue that student list products exacerbate racial inequality in college access. @RN5012 conceptualize student list products as "selection devices" [@RN4778] that enable colleges to select which prospective students they target by incorporating search filters (e.g., high school graduating class, state). @RN4786 defines "racialized inputs" as ostensibly race-neutral inputs that are systematically correlated with race because marginalized racial/ethnic groups have historically been excluded from the input [@RN4786]. @RN5012 argue that several frequently utilized student list filters (zip code, AP test score, SAT score) meet the criteria of racialized inputs. Using a national sample of high school students and using data from actual student lists purchased by public universities, @RN5012 show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects.

Because of the extent of residential segregation in the U.S., geographic borders are a commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887]. Student list products offer many geographic filters. Some geographic filters are based on known geographic borders (e.g., zip code, county, CBSA, state). College Board also created geographic borders that subsequently became filters in the *Student Search Service*, for example the Geodemographic Segment filters.[^14] This manuscript focuses on Geomarkets. Drawing from @RN4786, we argue that geomarkets satisfy the two criteria of racialized inputs. First, they are ostensibly race-neutral inputs in that neither @RN4982, nor subsequent promotional material [e.g., @RN1621], mention race. Second, Geomarkets may be systematically correlated with race. For example, Geomarket "CA07" is "City of Oakland" and "CA08" is "Alameda Country, excluding Oakland." A university might purchase prospect contact information by filtering on CA08 in combination with additional filters, such as year of high school graduation and SAT test score. This article extends @RN5012 by examining how the racial composition of actual student list purchases changes if universities filter on particular Geomarkets.

[^14]: For example, the geodemographic segment filters utilize cluster analysis to allocate each census tract and each high school into different categories based past college enrollment and other factors [@RN4565].

@@

> Imagine for the moment that you recruit for a college that draws most of its students from regional or national segments. Where would you concentrate your energies? Ideally, you would seek communities \[Geomarkets\] with a high proportion of students already predisposed toward institutions such as your own. The Market Segment Model would provide this information through segment percentages for the community in question. Further classification of students by social attributes allows you to identify a group for mailings or recruiting\ldots If you were to recruit in Boston, only about two out of every ten students with fewer than two attributes would likely listen, while slightly less than half of the students with two or more attributes would be receptive\ldots. Your efforts would surely be better directed toward \ldots Manchester, Hartford, and Fairfield County \ldots Indeed, in Fairfield County alone you could reach more than 40 percent of your "primary target" population -- that is, students with a greater than 75-percent probability of concentrating their college choices among institutions like your own.

>>>>

**Geomarket Filter in Student Search Service**. <!-- By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. --> A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the “enrollment funnel.” @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.[^11] Historically, dominant list vendors are college board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

@@

However, @RN5012 argue that student list products may exacerbate racial inequality in college access. The authors conceptualize student list products as "selection devices" [@RN4778] that enable colleges to select which prospective students they target by incorporating search filters (e.g., high school graduating class, state). @RN4786 defines "racialized inputs" as ostensibly race-neutral inputs that are systematically correlated with race because marginalized racial/ethnic groups have historically been excluded from the input [@RN4786]. @RN5012 argue that several frequently utilized student list filters (e.g., zip code, AP test score, SAT score) meet the criteria of racialized inputs. Using a national sample of high school students and using data from actual student lists purchased by public universities, @RN5012 show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects.

Because of the extent of residential segregation in the U.S., geographic borders are a commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887]. Student list products offer many geographic filters. Some geographic filters are based on known geographic borders (e.g., zip code, county, CBSA, state). College Board also created geographic borders that subsequently became filters in the *Student Search Service*, for example the Geodemographic Segment filters.[^14] This manuscript focuses on Geomarkets. In their sample, @RN5012 found that 11% of student lists purchased by public research universities filtered on Geomarkets. Drawing from @RN4786, we argue that geomarkets satisfy the two criteria of racialized inputs. First, they are ostensibly race-neutral inputs in that neither @RN4982 -- nor subsequent EPS promotional material [e.g., @RN1621] -- mention race. Second, Geomarkets may be systematically correlated with race. For example, Geomarket "CA07" is "City of Oakland" and "CA08" is "Alameda Country, excluding Oakland." A university might purchase prospect contact information by filtering on CA08 in combination with additional filters, such as year of high school graduation and SAT test score. This article extends @RN5012 by examining how the racial composition of actual student list purchases changes if universities filter on particular Geomarkets.

[^14]: Geodemographic segment filters utilize cluster analysis to allocate each census tract and each high school into different categories based past college enrollment and other factors [@RN4565].

@@

> Imagine for the moment that you recruit for a college that draws most of its students from regional or national segments. Where would you concentrate your energies? Ideally, you would seek communities \[Geomarkets\] with a high proportion of students already predisposed toward institutions such as your own. The Market Segment Model would provide this information through segment percentages for the community [Geomarket] in question. Further classification of students by social attributes allows you to identify a group for mailings or recruiting\ldots If you were to recruit in Boston, only about two out of every ten students with fewer than two attributes would likely listen, while slightly less than half of the students with two or more attributes would be receptive\ldots. Your efforts would surely be better directed toward \ldots Manchester, Hartford, and Fairfield County \ldots Indeed, in Fairfield County alone you could reach more than 40 percent of your "primary target" population -- that is, students with a greater than 75-percent probability of concentrating their college choices among institutions like your own.
